ADNI Data Analysis
========================================================
This code caclculates the slope of the PET volumes from 4 visits and explores the relationship between this slope and the probabilistic grey matter segmentation at baseline. 

Elizabeth Sweeney
June 15, 2014


Load required packages 
```{r}
library(oro.nifti)
library(RColorBrewer)
```

Set working directory 
```{r}
setwd('/dexter/disk1/smart/AD/ADNI/Greven_ADNI')
## setwd('/Users/elizabethsweeney/Desktop/Greven_ADNI')
```

Grab the subject ids 
```{r}
fileDir <- '/dexter/disk1/smart/AD/ADNI/Greven_ADNI'
files <- dir(fileDir, pattern = "*.nii.gz", full.names = TRUE)
subject.ids <- unlist(lapply(files, function(x) tail(unlist(strsplit(x, '/')), n=1)))
subject.ids <- unique(unlist(lapply(subject.ids, function(x) substr(x, 1, 10))))
```


Read in PET volumes and calculate the slope at each voxel.  The slope volume is then saved. 

```{r}
times <- matrix(c(0,  12, 36, 48),  4, 1)
intercept <- matrix(c(1,1,1,1), 4, 1) 
design <- cbind(intercept, times)
almost.beta <- solve(t(design)%*%design) %*% t(design)

for(i in 1:length(subject.ids)){
  pet1 <- readNIfTI(paste0(subject.ids[i], '_PET_coreg_to_MRI_MNI_1.nii.gz'), reorient = FALSE)[,,]
  pet2 <- readNIfTI(paste0(subject.ids[i], '_PET_coreg_to_MRI_MNI_2.nii.gz'), reorient = FALSE)[,,]
  pet3 <- readNIfTI(paste0(subject.ids[i], '_PET_coreg_to_MRI_MNI_3.nii.gz'), reorient = FALSE)[,,]
	pet4 <- readNIfTI(paste0(subject.ids[i], '_PET_coreg_to_MRI_MNI_4.nii.gz'), reorient = FALSE)[,,]

pet.image <- cbind(c(pet1), c(pet2), c(pet3), c(pet4))
pet.slope <- almost.beta%*%t(pet.image)
pet.slope.volume <- array(pet.slope[2,], dim = dim(pet1))

y.var <- apply(pet.image, 1, function(x) sd(x))
t.stat <- (4 - 2)* pet.slope[2,]/(solve(t(times)%*%times)*y.var) 
p.value <- dt(t.stat, df = 2)
p.value <- array(p.value , dim = dim(pet1))

writeNIfTI(p.value, paste0(subject.ids[i], '_pvalues'))

correction.size <- mean( c(sum(pet1 > .01, na.rm = TRUE) , sum(pet2 > .01, na.rm = TRUE), sum(pet3 > .01, na.rm = TRUE), sum(pet4 > .01, na.rm = TRUE)))

p.value.bonferoni.mask <- array(0, dim = dim(p.value))
p.value[is.na(p.value)] <- 10
p.value.bonferoni.mask[p.value  < .05/correction.size] <- 1
pet4[is.na(pet4)] <- 0
p.value.bonferoni.mask[pet4 < .01] <- 0


writeNIfTI(p.value.bonferoni.mask, paste0(subject.ids[i], '_bonfmask'))


writeNIfTI(pet.slope.volume, paste0(subject.ids[i], '_PETslope'))
print(i)
}
```

Explore relationship between PET slope and probabilistic grey matter segmentation in different ROIs

Hippocampus
```{r}
grey.matter <- c()
pet.slopes <- c()

ROI <- readNIfTI('/dexter/disk1/smart/AD/ADNI/Greven_ADNI/AAL_masks/Hippocampus.nii.gz', reorient = FALSE)[,,]

for(i in 1:length(subject.ids)){
  PET <- readNIfTI(paste0(subject.ids[i], '_PETslope'), reorient = FALSE)[,,]
  GM <- readNIfTI(paste0(subject.ids[i], '_MRI_segm_MNI_1.nii.gz'), reorient = FALSE)[,,]
  grey.matter <- c(grey.matter, GM[ROI == 1]) 
  pet.slopes <- c(pet.slopes, PET[ROI == 1])
  print(i) 
  }

pdf('Hippo.pdf')
down.samp <- sample(1:length(grey.matter), 5000, replace = FALSE)
smooth <- loess.smooth(grey.matter, pet.slopes) 
plot(pet.slopes[down.samp] ~ grey.matter[down.samp], pch = 20, main = 'Hippocampus', ylab = 'PET slope', xlab = c('Baseline Grey Matter'))
lines(smooth, lwd = 3, col = 'red')
dev.off()

pdf('SubjectsHippo.pdf')
down.samp <- sample(1:length(grey.matter), 5000, replace = FALSE)
plot(pet.slopes[down.samp] ~ grey.matter[down.samp], pch = 20, main = 'Hippocampus by Subject', ylab = 'PET slope', xlab = c('Baseline Grey Matter'))


colors.mine <- rep(brewer.pal(9, 'Set1'), 12)

for(i in 1:length(subject.ids)){
  PET <- readNIfTI(paste0(subject.ids[i], '_PETslope'), reorient = FALSE)[,,]
  GM <- readNIfTI(paste0(subject.ids[i], '_MRI_segm_MNI_1.nii.gz'), reorient =   FALSE)[,,] 
  subj.grey.matter <- GM[ROI == 1]
  subj.pet.slopes <- PET[ROI == 1]
  smooth <- loess.smooth(subj.grey.matter, subj.pet.slopes) 
  lines(smooth, lwd = 1, col = colors.mine[i])
  print(i)
}
dev.off()


```

Make PLOT of the PET slope volume for one subject 
```{r}

PET <- readNIfTI(paste0(subject.ids[1], '_PETslope'), reorient = FALSE)[,,]
PET[is.na(PET)] <- 0

pdf('petslopes.pdf')
orthographic(PET, col = hotmetal())
dev.off()  

mean.pet <- array(0, dim = dim(PET))
for(i in 1:length(subject.ids)){
  PET <- readNIfTI(paste0(subject.ids[i], '_PETslope'), reorient = FALSE)[,,]
  PET[is.na(PET)] <- 0
  mean.pet <- ((i-1)*mean.pet + PET)/i
  }

var.pet <- array(0, dim = dim(PET))
for(i in 1:length(subject.ids)){
  PET <- readNIfTI(paste0(subject.ids[i], '_PETslope'), reorient = FALSE)[,,]
  PET[is.na(PET)] <- 0
  var.pet <- ((i-1)*var.pet + (PET - mean.pet)^2)/i
  }

setwd('/dexter/disk1/smart/AD/ADNI/Greven_ADNI/Output')

writeNIfTI(mean.pet,'meanPETslope')
writeNIfTI(var.pet,'varPETslope')

```